Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses

### 2014-10-27 paper_v1.Rnw
some updates to s1

\begin{abstract}
Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from 20879 online learners in 18 courses. Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. Study 2 is a case study to develop a deeper understanding of attrition in MOOCs by conducting a case study. Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey. (abstract needs work...)
\end{abstract}

@@

Educational environments has become increasingly diverse. Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost and by accommodating people with less flexible schedules \cite{goldrick2010challenges}. Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (``correspondence education''), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

@@

@@

% \begin{description}
%   \item[H1] prior experience -- technical difficulties
%   \item[H2] demographics (education) -- course difficulty
% \end{description}

@@

% Which courses did we use? What data
Seventeen MOOCs on topcis in a variety of disciplines were selected for investigation: four on topics in the humanities, two in engineering, three in computer science, two runs of the same course in business, and six in mathematics and statistics. For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment. Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). Table \ref{tab:study1_courses_summary} provides the following summary statistics for each course: the number of enrolled learners, the number of learners who completed the course survey and the feedback surveys, age, gender, and education level, and the proportion of learners who self-reported that they stopped participating in the course. Note that age, gender, and education level statistics were computed using only the subset of learners who took both the initial and feedback surveys. We also computed the proprtions of learners who viewed at least 30, 50, and 80\% of the course videos.

@@

\begin{table*}
\label{tab:study1_courses_summary}
\caption{Summary statistics of 14 MOOCs}
\small
\center
\begin{tabular}{llcccccccccc}
\toprule
%  & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & \specialcellcentered{\ensuremath{>} 30\% \\ videos} & \specialcellcentered{\ensuremath{>} 50\% \\ videos} & \specialcellcentered{\ensuremath{>} 80\% \\ videos} & \specialcellcentered{Self-Rep. \\ Dropout} \\
& & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
& Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\midrule
C1 & Business & 61,096 & 9,943 & 2,367 & 42\% & 39 (30,49) & n/a & 39\% & 20\% & 7\% & 10\% \\
C2 & Business & 41,081 & 6,544 & 883 & 40\% & 39 (30,49) & 97\% & 32\% & 13\% & 2\% & 9\% \\
C3 & CS & 27,104 & 6,104 & 2,854 & 35\% & 30 (22,41) & 74\% & 67\% & 40\% & 7\% & 2\% \\
C4 & CS & 56,154 & 3,406 & 1,123 & 11\% & 33 (27,42) & 91\% & 34\% & 16\% & 11\% & 21\% \\
C5 & CS & 57,636 & 5738 & 402 & 17\% & 33 (26,43) & 91\% & 66\% & 55\% & 28\% & 14\% \\
C6 & Engineering & 16,220 & 2,540 & 269 & 13\% & 33 (27,45) & 92\% & 62\% & 49\% & 17\% & 15\% \\
C7 & Enginerring & 7,178 & 1,312 & 202 & 12\% & 27 (22,39) & 85\% & 51\% & 18\% & 2\% & 7\% \\
C8 & Humanities & 7,634 & 3,631 & 152 & 92\% & 33 (26,45) & 93\% & 45\% & 11\% & 0\% & 7\% \\
C9 & Humanities & 6,475 & 3,315 & 326 & 90\% & 32 (27,47) & 91\% & 69\% & 50\% & 17\% & 16\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570 & 52\% & 39 (29,52) & 89\% & 84\% & 74\% & 21\% & 3\% \\
C11 & Humanities & 630 & 167 & 80 & 50\% & 30 (27,40) & 97\% & 74\% & 55\% & 16\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862 & 33\% & 41 (29,55) & 88\% & 27\% & 10\% & 1\% & 31\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & 58\% & 28 (16,46) & n/a & 69\% & 38\% & 8\% & 4\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363 & 9\% & 31 (26,38) & 99\% & 61\% & 45\% & 25\% & 14\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & 20\% & 35 (29,45) & 98\% & 78\% & 67\% & 34\% & 8\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637 & 82\% & 43 (34,51) & n/a & 64\% & 36\% & 5\% & 1\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843 & n/a & n/a & n/a & 70\% & 54\% & 14\% & 5\% \\
\bottomrule
\end{tabular}
\end{table*}

% Proportions for each dropout reason for each course
We are interested in the proportions of respondents that reported each reason for dropout across the courses. These proportions are shown in Figure \ref{fig:prop-reason}. The proprtion for each reason was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it aplies or that it does not apply). We also computed the median and interquartile range of the proprtions across all courses for each reason.

%Figure: proportions for each dropout reason for each course
<<prop-reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
57,39,13,48,42,51,33,60,65,42,18,52,21,68,39,28,68,
21,23,25,14,13,19,8,0,0,23,9,12,20,20,13,28,12,
32,18,25,28,32,28,25,11,20,27,27,19,19,22,18,0,11,
15,14,8,16,13,12,0,0,9,10,9,9,19,12,7,14,14,
21,3,18,12,17,24,25,30,30,10,36,13,20,25,12,0,21,
69,58,38,76,80,66,66,55,69,67,27,71,28,64,54,0,71,
12,7,27,8,4,18,41,11,6,10,27,10,24,14,11,28,7,
18,7,26,17,19,16,41,10,9,25,18,21,27,25,20,28,12,
13,12,10,19,19,21,0,11,7,13,9,9,11,16,9,14,19,
16,5,22,17,14,21,16,11,7,18,20,16,23,20,16,16,12,
14,10,10,23,24,16,8,11,2,13,18,19,10,25,17,0,29,
26,8,11,27,23,39,33,11,24,11,27,27,11,35,15,14,41,
14,9,18,10,10,18,25,22,4,5,18,7,20,22,14,14,25)

@@

The feedback survey for each course asked the learners, among other questions, whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop. For each reason, the student was asked to select one of two options: "Did influence my decision to stop" and "Did NOT influence my decision to stop". Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reportinga reason and those who express that the reason did not apply to them. We are interested in examining the strength of correlations between these reasons. We computed the correlation matrix for the reasons for each course separately, and then linearly combined these matrices to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}.

@@

In the initial surveys, learners were asked about their reasons for enrollment, and they were asked to mark whether or not each of a set of enrollment reasons \cite{} applies to them. For each course, we computed the correlation matrix of the reasons for dropout with reasons for enrollment. These matrices were then linearly combined to give the matrix shown in Table \ref{tab:study1_rfd_ei_corr}.

@@

% Proportion videos watched for each disengagement reason
We are also interested in the correlations between reported reasons for dropout and certain learner behaviors. For each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes). We then computed the matrix of correlations between reasons for dropout and these variables for each course. These matrices were averaged and the result is shown in Table [TableToBeAdded].

\subsection{Results}
Study 1 shows that the two most commonly reported reasons for dropout are related to one's ability to commit time to the course. The correlation between the two reasons is $r=0.32$. A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.

@@

\begin{table*}
\label{tab:rfd_ei_strongest_correlations}
\caption{Strongest observed correlations ($r >= 0.15) between reasons for disengagement and enrollment intentions}
\center
\small
\begin{tabular}{llr}
\toprule
 Enrollment intention & Dropout reason & $r$ \\
\midrule
Take with others & Course not challenging enough & 0.24 \\
Take with others & Too advanced for my background & 0.20 \\
Experience an online course & I was only exploring the course & 0.17 \\
Improve my English & Technical difficulties & 0.17 \\
Personal growth & Could not keep up with deadlines & 0.16 \\
Take with others & Credentials not valuable enough for me & 0.16 \\
Take with others & Technical difficulties & 0.15 \\
Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Discussion}
Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward. Expectency-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain. In other words, a learner might disengage from the course if she gets occupied by other higher priority tasks. Because learners can join or leave MOOCs freely. Joining a MOOC does not necessarily imply that a learner is capable of allocating the time for it according to the rational framework that expectancy value theory predicts. Moreover, previous research on volitional control suggests that learners allocate time to the next highest priority task, but could rather get distracted by lower priority activities that promise nearer gratification. So what does lack of time for the course mean? In study 2, we attempt to investigate various possible dropout reasons that learners might report as "lack of time".

[TBC]

>>>>

\begin{abstract}
% Persistence and attrition in MOOCs are systematically analyzed using self-report and behavioral data collected from 20879 online learners in 18 courses. Study 1 offers insights into reasons for disengaging from MOOCs and explores relationships with prior behavior and reported intentions. Study 2 is a case study to develop a deeper understanding of attrition in MOOCs by conducting a case study. Targeting online learners who were predicted likely dropouts in a particular course were invited to provide feedback via a survey. (abstract needs work...)
Abstract goes here.
\end{abstract}

@@

Educational environments have become increasingly diverse. Traditional schools and universities have a characteristically rigid structure, including instructor-defined---even nationally agreed---syllabi, fixed time schedules, entry requirements, and material costs to enter and exit. Novel institutional structures have been developed to overcome particular constraints. Community colleges, for instance, were created in an attempt to democratize education by offering instruction at a lower cost and by accommodating people with less flexible schedules \cite{goldrick2010challenges}. Distance learning programs intended to deliver education in remote parts of the world and for people who simply could not attend in-person classes. Course materials, including assessments, were delivered through mail (``correspondence education''), radio, television, and eventually the Internet, thereby addressing geographical and time related constraints of traditional instruction \cite{moore1996distance}.

@@

\begin{table*}[th]
\label{tab:s1sum}
\caption{Summary statistics of learner demographics and behavior in 14 MOOCs}
\small
\center
\begin{tabular}{llccccccccccc}
% & & & & & & & & $>$ 30\% & $>$ 50\% & $>$ 80\% & Self-Rep. \\
% & Topic Area & Enrollment & Course survey & Feedback survey & Female & Age & Education & videos & videos & videos & Dropout \\
\toprule
 &            &          & \multicolumn{3}{c}{Survey Responses} & & & & & \multicolumn{3}{c}{Videos Watched} \\
 \cmidrule(r){4-6} \cmidrule(r){11-13}
 & Topic Area & Enrolled & Initial & Feedback & Both & Female & Age$^1$   & Education$^2$ & Dropout$^3$ & $>$ 30\% & $>$ 50\% & $>$ 80\% \\
\midrule
C1 & Business & 61,096   & 9,943  & 2,367    & N    & 42\%   & 39 (30,49) & n.a.       & 10\% & 39\% & 20\% & 7\% \\
C2 & Business & 41,081   & 6,544  & 883      & N    & 40\%   & 39 (30,49) & 97\%       & 9\%  & 32\% & 13\% & 2\% \\
C3 & CS       & 27,104   & 6,104 & 2,854     & N    & 35\%   & 30 (22,41) & 74\%       & 2\%  & 67\% & 40\% & 7\% \\
C4 & CS       & 56,154   & 3,406 & 1,123     & N    & 11\%   & 33 (27,42) & 91\%       & 21\% & 34\% & 16\% & 11\% \\
C5 & CS       & 57,636   & 5738  & 402       & N    & 17\%   & 33 (26,43) & 91\%       & 14\% & 66\% & 55\% & 28\% \\
C6 & Engineering & 16,220 & 2,540 & 269      & N    & 13\%   & 33 (27,45) & 92\%       & 15\% & 62\% & 49\% & 17\% \\
C7 & Enginerring & 7,178 & 1,312 & 202       & N    & 12\%   & 27 (22,39) & 85\%       & 7\%  & 51\% & 18\% & 2\% \\
C8 & Humanities & 7,634 & 3,631  & 152       & N    & 92\%   & 33 (26,45) & 93\%       & 7\%  & 45\% & 11\% & 0\% \\
C9 & Humanities & 6,475 & 3,315  & 326       & N    & 90\%   & 32 (27,47) & 91\%       & 16\% & 69\% & 50\% & 17\% \\
C10 & Humanities & 9,903 & 4,227 & 1,570     & N    & 52\%   & 39 (29,52) & 89\%       & 3\%  & 84\% & 74\% & 21\% \\
C11 & Humanities & 630 & 167     & 80        & N    & 50\%   & 30 (27,40) & 97\%       & 16\% & 74\% & 55\% & 16\% \\
C12 & Math \& Stats & 49,317 & 3,090 & 862   & N    & 33\%   & 41 (29,55) & 88\%       & 31\% & 27\% & 10\% & 1\% \\
C13 & Math \& Stats & 32,006 & 26,142 & 5,336 & N   & 58\%   & 28 (16,46) & n.a.       & 4\%  & 69\% & 38\% & 8\% \\
C14 & Math \& Stats & 16,857 & 2,892 & 363   & N    & 9\%    & 31 (26,38) & 99\%       & 14\% & 61\% & 45\% & 25\% \\
C15 & Math \& Stats & 37,017 & 11,629 & 1,854 & N   & 20\%   & 35 (29,45) & 98\%       & 8\%  & 78\% & 67\% & 34\% \\
C16 & Math \& Stats & 27,236 & 2,711 & 637   & N    & 82\%   & 43 (34,51) & n.a.       & 1\%  & 64\% & 36\% & 5\% \\
C17 & Math \& Stats & 13,132 & 3,629 & 843   & N    & n.a.   & n.a.       & n.a.       & 5\%  & 70\% & 54\% & 14\% \\
\bottomrule
\multicolumn{13}{l}{$^1$ mean (SD)  $^2$ bachelors' or higher degree  $^3$ self-reported in the feedback survey}
\end{tabular}
\end{table*}

@@

The present study extends the existing literature on persistence with a representative quantitative treatment of attrition in MOOCs. We address the research questions developed in the previous section with a combination of self-report and behavioral data from seventeen diverse MOOCs. Going beyond specific research questions, this observational analysis is intended to inform our understanding of attrition in MOOCs to better support learners in these environments. This can be achieved through general changes, for instance, to the course format; or by targeted interventions that support a particular group of learners, such as those who exhibit low levels of goal striving (cf. Study 2).

@@

Seventeen MOOCs on topics in a variety of disciplines were selected for investigation (see Table \ref{tab:s1sum}). For each course, we merged data from the initial course survey (distributed in the first weeks of the course), the feedback surveys (distributed in the last weeks of the course), and behavioral data collected in the learning environment. 
% Only those learners who completed the feedback survey were included in the analyses (some did not complete the initial course survey). 
Table \ref{tab:s1sum} provides summary statistics for enrollment, survey response rates, demographics, self-reported dropout, and behavioral indicators for each course. Self-reported dropout was measured in the feedback survey by asking learners if they stopped participating in the course. The three behavioral indicators are the proportions of learners who viewed at least 30, 50, and 80\% of the lecture videos. Information about age, gender, and education level were only available for learners who completed the initial course survey.

@@

\subsubsection{Measures}

Learners reported reasons for enrolling in the course (using the OLEI scale \cite{kizilcec2015motivation} on the course survey) and leaving the course (on the feedback survey) on a grid of binary-choice items. The feedback survey for asked learners whether or not they have stopped participating before the end of the course, and if they did, what reasons influenced their decision to stop. For each reason, the student was asked to select one of two options: ``Did influence my decision to stop" and ``Did NOT influence my decision to stop". Compared to a question where the student only checks a box if the reason applies, this design aims to distinguish between respondents who skip reporting reasons and those who express that the reason did not apply to them. We also computed the fraction of videos in the course each learner had watched to relate self-report measures to learner behavior.

We first report the proportion for each reason for dropping out. This was computed as the number of respondents who reported that the reason applied to them divided by the total number of those who responded to the question (reporting that it applies or that it does not apply). We also computed the median and interquartile range of the proprtions across all courses for each reason. The resulting proportions are shown in Figure \ref{fig:prop-reason}. Second, we performed three correlational analyses were performed to investigate associations between these measures.

We examined the strength of correlations between reasons for dropout. The correlation matrix for the reasons were computed separately for each course, and then linearly combined to yield the average correlation matrix shown in Table \ref{tab:study1_rfd_corr}. Following the same procedure, we computed the correlation matrix between reasons to enroll and reasons for attrition. Furthermore, we computed correlations between reported reasons for dropout and certain learner behaviors. Finally, for each learner, we computed the fraction of course videos visited (the learner started playing the video) and fraction watched (the learner stayed with the video for 3 minutes, or for the whole video duration if shorter than 3 minutes). We then computed the matrix of correlations between reasons for dropout and these variables for each course. These matrices were averaged and the result is shown in Table [TableToBeAdded].

<<prop-reason, echo=FALSE, fig.env='figure*', fig.pos='ht', fig.height=4, fig.width=9, fig.cap='Proportion of learners who indicated each reason for attrition across 17 MOOCs'>>=
values = c(18,20,10,22,19,21,8,11,13,13,9,7,14,18,11,14,19,
57,39,13,48,42,51,33,60,65,42,18,52,21,68,39,28,68,
21,23,25,14,13,19,8,0,0,23,9,12,20,20,13,28,12,
32,18,25,28,32,28,25,11,20,27,27,19,19,22,18,0,11,
15,14,8,16,13,12,0,0,9,10,9,9,19,12,7,14,14,
21,3,18,12,17,24,25,30,30,10,36,13,20,25,12,0,21,
69,58,38,76,80,66,66,55,69,67,27,71,28,64,54,0,71,
12,7,27,8,4,18,41,11,6,10,27,10,24,14,11,28,7,
18,7,26,17,19,16,41,10,9,25,18,21,27,25,20,28,12,
13,12,10,19,19,21,0,11,7,13,9,9,11,16,9,14,19,
16,5,22,17,14,21,16,11,7,18,20,16,23,20,16,16,12,
14,10,10,23,24,16,8,11,2,13,18,19,10,25,17,0,29,
26,8,11,27,23,39,33,11,24,11,27,27,11,35,15,14,41,
14,9,18,10,10,18,25,22,4,5,18,7,20,22,14,14,25)

@@

@@

@@

(Note: this section is work in progress, we will go through each RQ and present results that address them here.)

Table \ref{tab:study1_rfd_ei_corr}.

How do prior learner characteristics (demographics, geographic location, intentions and motivations, and previous online courses) influence learners' likelihood of dropping out of a MOOC? (RQ1)

What proportion of learners who dropped out were satisfied with their achievements in the course? (RQ2)

What reasons do learners report for dropping out of a MOOC? (RQ3a)

 How do reasons for dropping out vary across prior learner characteristics (cf. RQ1)? (RQ3b)

How does behavior in MOOCs vary by reasons for dropping out? (RQ4)

The two most commonly reported reasons for dropout are related to one's ability to commit time to the course. The correlation between the two reasons is $r=0.32$. A more detailed analysis we conditioned on responses to these two reasons revealed that 30\% of respondents who reported at least one of these two time reasons reported no other reasons for dropout, while 24\% reported one other reason and 46\% reported more than one.

@@

% \begin{table*}
% \label{tab:rfd_ei_strongest_correlations}
% \caption{Strongest observed correlations ($r >= 0.15$) between reasons for disengagement and enrollment intentions}
% \center
% \small
% \begin{tabular}{llr}
% \toprule
%  Enrollment intention & Dropout reason & $r$ \\
% \midrule
% Take with others & Course not challenging enough & 0.24 \\
% Take with others & Too advanced for my background & 0.20 \\
% Experience an online course & I was only exploring the course & 0.17 \\
% Improve my English & Technical difficulties & 0.17 \\
% Personal growth & Could not keep up with deadlines & 0.16 \\
% Take with others & Credentials not valuable enough for me & 0.16 \\
% Take with others & Technical difficulties & 0.15 \\
% Prestigious uni/prof & Could not keep up with deadlines & 0.15 \\
% \bottomrule
% \end{tabular}
% \end{table*}

(Note: this section is also work in progress.)

Even though study 1 accentuates ``time'' as a reported reason for dropout, interpreting what time means is not straight forward. Expectency-value theory models the learner as a rational agent who allocates time to different activities to maximize subjective gain. In other words, a learner might disengage from the course if she gets occupied by other higher priority tasks. Because learners can join or leave MOOCs freely. Joining a MOOC does not necessarily imply that a learner is capable of allocating the time for it according to the rational framework that expectancy value theory predicts. Moreover, previous research on volitional control suggests that learners allocate time to the next highest priority task, but could rather get distracted by lower priority activities that promise nearer gratification. So what does lack of time for the course mean? In study 2, we attempt to investigate various possible dropout reasons that learners might report as ``lack of time".
